Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-07 csf-2017/paper.tex
A pass over the introduction.

\usepackage{cite}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{thmtools}
\usepackage{stmaryrd}
\usepackage{natbib}
\usepackage{url}
\usepackage{array}
\usepackage{arydshln}
\usepackage{ifthen}
\usepackage{ifpdf}
\usepackage{verbatim}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{hyperref}
\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  keepspaces=true,
  mathescape
}

@@

\section{Introduction}
\label{sec:introduction}
\todo[inline]{See the latex comments for the next paragraphs!}
% Possible high-level intros:
% 0) Arms race compilers
%
% 1) Security guarantees vs obstacles / arms race
\begin{comment}
  Security on processors is today subject to an arms race between
  processor manufacturers and hackers. On the hacker side, a constant
  effort is put into exploiting current processor designs and
  circumventing the security measures set in place to prevent known
  exploits. On the processor manufacturer side, new security measures
  are put in place to prevent new exploits. The arms race would come
  to a grinding halt if the security meassures provided more than
  obstacles and actually provided low-level security guarantees. One
  kind of low-level machines that provide low-level security is a
  capability machine.
\end{comment}
% 2) Fully abstract compilation
High-level languages provide guarantees such as encapsulation of local
state and well-bracketedness of function calls. These guarantees are
often relied upon when reasoning about the correctness and security of
a program. High-level programs are compiled to machine code, so in
order to rely on the correctness and security guarantees provided by
the high-level language, it needs to be proven that the translation
preserves these properties. Translated programs often interact with
machine code that is not translated from the same high-level language,
so in order to preserve the guarantees from the high-level languages,
the low-level machine needs to provide some kind of security
guarantees. In other words, a target of secure compilation needs to
provide some security guarantees. Modern processors provide many
security meassures, but very little in terms of guarantees. A machine
that provides some means of security guarantees is a capability machine.

% 3) Memory access-control
% I don't like this angle and can't find a good way to sell it.

% 4) History
% Proposals throughout history.
% Lack of formal account of properties-guarantees.

%%% a)
% Current low-level protection coarse grained memory protection or
% properties of high-level languages are not really enforced?

% Capability machines offer fine-grained memory protection
% Too technical
A capability machine is a low-level machine that offers fine-grained
memory protection.
% Capability machines and unforgeable tokens for memory access. 
This is done by introducing unforgeable tokens to
the machine. These tokens are called capabillities and grant a kind of
authority and a range of authority. 
% Dynamic checks encapsulation, formal model

 We will refer to the capabilities
that grant a combination of read, write, or execute permission as memory
capabilities. 
 Capabilities with the necessary permissions must be
presented every time an instruction that manipulates the memory is
executed.

Memory capabilities provide memory protection, but they do not provide
a way to set up security domains. When you jump to an untrusted piece
of code that is supposed to jump back to you, then you have to either
revoke all your capabilities or pass to the piece of code you do not
trust. This short coming can be overcome by adding something like the
enter capability from the M-Machine\todo{add reference}. The enter
capability is an opaque capability which can only be used for a
jump. When jumped to it grants read and execute permission which
allows the code that is now executing to read capabilities stored in
memory. 
The CHERI processor's ccall achieves a similar property\todo{Add reference}.

Capabilities are irrevocable, so when we pass a capability has to
an untrusted program, then we have to assume that the program keeps it
around indefenitely. This means that we cannot reuse the piece of
memory that the capability governs essentially creating a memory
leak. The CHERI processor has a special kind of capabilities called
\emph{local capabilities}\todo{reference?}\ that introduces a simple
kind of temporal information control which allows for a simple kind of
capability revocation. This is achieved by adding a tag to every
capability that marks it as either local or global. Local capabilities
can only be written through a capability with a new permission called
\emph{permit write local}.
% More about local capabilities.

With just memory capabilities, enter capabailities, and local
capabilities, the capability machine is very powerful and can
enforce properties that high-level languages promise. In most
high-level languages, we expect local state to be private and
encapsulated. The memory capabilities make sure that local state
cannot be accessed without a capability to do so. With the enter
capabaility, we can regain control of local state after passing
control to an untrusted piece of code which gives us encapsulation
when control is passed to an untrusted program.

Many high-level languages guarantee that function function calls are
well-bracketed. This can easily be achieved by having a trusted call
stack. On a capability machine, it is possible to enforce well
barcketedness without a trusted stack. When ensuring well-bracketed
the two main challenges calls are 1) preventing storage of return
pointers and 2) encapsulation of the local stack frame. If we don't
have the first property, then an adversary can store the return
capability in a call and jump to it in a later call. If the local
stack frame is not properly encapsulated, then an adversary can break
the well-bracketedness by jumping to a different calls return
capability.

In order to deal with the two challenges, we make sure that all return
capabilities are local and that there are no global capabilities with
permit write local permission. This makes sure that there is no way to
save a return capability in persistent local state.\lau{this does not really make
sense as the local stack frame can be seen as local state} It is,
however, to limiting to not provide any means to store the return
capabilities as a piece of code may need to store a return capability
before jumping to an adversary. It is therefore needed to provide a
local capability with permit write local permission which can be used
as a stack. The stack should now be the only place one can store the
return capability, but it should also be the only place one can store
the stack capability.

Using local return capabilities and putting a stack abstraction on a
local capability permit write local capability is a good start, but as
it turns out, it is not wnough. An adversary could do the following 
\begin{enumerate}
\item In a first invocation, fill the stack frame with copies of the
  current stack capability.
\item In a later call, the adversary may be able to load the stack
  capability from the previous call.
\item If we have used a larger part of the stack than in the first
  call, then the adversary has access to part of our local stack frame.
\end{enumerate}

The above attack shows that it is tricky to ensure
well-bracketedness. It raises the question whether it is possible to
do so it is completely watertight. The attack utilized the stack, so
it is necessary to have a trusted stack or can we do without? In
general it raises the question, how do we reason about capability
machines and local capabilities, so we are sure that what ever calling
scheme we come up with actually provably ensures well-bracketed calls.

In the following paper, we present the following contributions
\begin{enumerate}
\item 
\end{enumerate}

>>>>

\usepackage{cite}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{thmtools}
\usepackage{stmaryrd}
\usepackage{natbib}
\usepackage{url}
\usepackage{array}
\usepackage{arydshln}
\usepackage{ifthen}
\usepackage{ifpdf}
\usepackage{verbatim}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{natbib}
\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  keepspaces=true,
  mathescape
}

@@

\section{Introduction}
\label{sec:introduction}
\todo[inline]{See the latex comments for the next paragraphs!}

Compromising software security is often based on attacks that break programming
language properties relied on by software authors, such as control flow
integrity, local state encapsulation etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control flow integrity and
local state encapsulation necessarily rely on techniques like machine code
rewriting (e.g.~\cite{wahbe_efficient_1993,abadi_control-flow_2005}) or
verification (e.g.~\cite{morrisett_system_1999}), virtual machines with a native
stack (e.g.~the JVM~\cite{lindholm_java_2014}) or
randomization~\cite{forrest_building_1997}. Such techniques for retrofitting
protection techniques to commodity hardware have disadvantages in terms of
performance, complexity and/or security.

\emph{Capability machines} are a type of processors, intended to remediate these
limitations by presenting a better security model at the hardware
level~\cite{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740,Carter:1994:HSF:195473.195579}.\dominique{todo:
  more?} Such hardware tags every word (in the register file and in memory) to
enforce a strict separation between numbers and capabilities, a kind of pointers
carrying some kind of authority. Memory capabilities carry the authority to read
and/or write to a range of memory locations. Additionally, the details vary but
all capability machines offer some form of \emph{object capabilities}, which
represent the authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g. the M-Machine's enter capabilities or CHERI's
sealed code/data pairs).

Unlike for commodity processors, the security primitives on capability machines
lend themselves well to the enforcement of properties like local state
encapsulation and control-flow integrity. Potentially, they will enable new
compilation schemes that enforce such properties in a way that is efficient but
also 100\% watertight (ideally evidenced by a mathematical proof, guaranteeing
that we do not end up in a new attack-defense arms race). However, there is
still a lot that needs to happen to attain this goal. For example, it is quite
non-trivial to devise a compilation scheme adapted to the details of a specific
language's notion of encapsulation (e.g. encapsulation of private member
variables in OO languages often behaves quite differently than encapsulation of
private state in ML-like languages). And even if such a scheme were defined, a
formal proof depends on a formalisation of the encapsulation provided by the
capability machine at hand.

A similar problem is the enforcement of control-flow integrity on capability
machines. An interesting approach is taken in CheriBSD. They drop the standard
contiguous C stack and split it into a central, trusted shadow stack, managed by
trusted call and return instructions\footnote{These call and return instructions
  are implemented by CheriBSD and special support from the CHERI processor makes
  them very fast to invoke.}, and disjoint, per-component
stacks~\citep{Watson2015Cheri}. For preventing illegal use of stack references,
the approach relies on \emph{local capabilities}, a type of capabilities offered
by CHERI that can be passed over a function invocation \emph{temporarily}, for
the duration of the invocation, and revoked afterwards. However, a lot of
details remain to be filled in (e.g. how to combine a trusted shadow stack with
(potentially) cross-domain function pointers in a scalable way?) and there is no
argument that this enforcement of CFI is watertight.

In this paper, we make progress towards the goal of efficient, provably
watertight enforcement of encapsulation and control-flow integrity on capability
machines.  Specifically, we make the following contributions:
\begin{itemize}
\item We formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  integrity and encapsulation of local state on the stack. The discipline relies
  solely on local capabilities (used to represent stack and return pointers) and
  does not require OS support in the form of a trusted shadow stack or special
  call and return instructions. It also does not require separate per-component
  stacks and instead works with a single, fairly standard contiguous stack. It
  supports higher-order cross-component calls (e.g. (potentially)
  cross-component function pointers) and can be efficient assuming one
  additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem (capability-safety) that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed logical
  relation with recursively defined Kripke worlds. It is inspired by one used by
  \cite{Devriese:2016ObjCap} for capability-safety in a JavaScript-like lambda
  calculus, but has two novel technical characteristics. The natural CPS style
  of assembly code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal closure that
  has been previously used for lambda calculi. Perhaps more surprisingly, we
  show that the special nature of local capabilities can be elegantly expressed
  using a variant of an idea that has been used by
  \cite{Dreyer:2010:IHS:1863543.1863566} to express well-bracketedness
  guarantees in a lambda calculus: Kripke worlds with a weak and a strong future
  world relations. Interestingly, the two future world relations generally work
  very similarly to Dreyer et al.'s, except that they turn up in quite a
  different place in the logical relation.
\item We demonstrate the power of our results by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow integrity guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability-safety and our stack management discipline.
\end{itemize}

% % Possible high-level intros:
% % 0) Arms race compilers
% %
% % 1) Security guarantees vs obstacles / arms race
% \begin{comment}
%   Security on processors is today subject to an arms race between
%   processor manufacturers and hackers. On the hacker side, a constant
%   effort is put into exploiting current processor designs and
%   circumventing the security measures set in place to prevent known
%   exploits. On the processor manufacturer side, new security measures
%   are put in place to prevent new exploits. The arms race would come
%   to a grinding halt if the security meassures provided more than
%   obstacles and actually provided low-level security guarantees. One
%   kind of low-level machines that provide low-level security is a
%   capability machine.
% \end{comment}
% % 2) Fully abstract compilation
% High-level languages provide guarantees such as encapsulation of local
% state and well-bracketedness of function calls. These guarantees are
% often relied upon when reasoning about the correctness and security of
% a program. High-level programs are compiled to machine code, so in
% order to rely on the correctness and security guarantees provided by
% the high-level language, it needs to be proven that the translation
% preserves these properties. Translated programs often interact with
% machine code that is not translated from the same high-level language,
% so in order to preserve the guarantees from the high-level languages,
% the low-level machine needs to provide some kind of security
% guarantees. In other words, a target of secure compilation needs to
% provide some security guarantees. Modern processors provide many
% security meassures, but very little in terms of guarantees. A machine
% that provides some means of security guarantees is a capability machine.

% % 3) Memory access-control
% % I don't like this angle and can't find a good way to sell it.

% % 4) History
% % Proposals throughout history.
% % Lack of formal account of properties-guarantees.

% %%% a)
% % Current low-level protection coarse grained memory protection or
% % properties of high-level languages are not really enforced?

% % Capability machines offer fine-grained memory protection
% % Too technical
% A capability machine is a low-level machine that offers fine-grained
% memory protection.
% % Capability machines and unforgeable tokens for memory access. 
% This is done by introducing unforgeable tokens to
% the machine. These tokens are called capabillities and grant a kind of
% authority and a range of authority. 
% % Dynamic checks encapsulation, formal model

%  We will refer to the capabilities
% that grant a combination of read, write, or execute permission as memory
% capabilities. 
%  Capabilities with the necessary permissions must be
% presented every time an instruction that manipulates the memory is
% executed.

% Memory capabilities provide memory protection, but they do not provide
% a way to set up security domains. When you jump to an untrusted piece
% of code that is supposed to jump back to you, then you have to either
% revoke all your capabilities or pass to the piece of code you do not
% trust. This short coming can be overcome by adding something like the
% enter capability from the M-Machine\todo{add reference}. The enter
% capability is an opaque capability which can only be used for a
% jump. When jumped to it grants read and execute permission which
% allows the code that is now executing to read capabilities stored in
% memory. 
% The CHERI processor's ccall achieves a similar property\todo{Add reference}.

% Capabilities are irrevocable, so when we pass a capability has to
% an untrusted program, then we have to assume that the program keeps it
% around indefenitely. This means that we cannot reuse the piece of
% memory that the capability governs essentially creating a memory
% leak. The CHERI processor has a special kind of capabilities called
% \emph{local capabilities}\todo{reference?}\ that introduces a simple
% kind of temporal information control which allows for a simple kind of
% capability revocation. This is achieved by adding a tag to every
% capability that marks it as either local or global. Local capabilities
% can only be written through a capability with a new permission called
% \emph{permit write local}.
% % More about local capabilities.

% With just memory capabilities, enter capabailities, and local
% capabilities, the capability machine is very powerful and can
% enforce properties that high-level languages promise. In most
% high-level languages, we expect local state to be private and
% encapsulated. The memory capabilities make sure that local state
% cannot be accessed without a capability to do so. With the enter
% capabaility, we can regain control of local state after passing
% control to an untrusted piece of code which gives us encapsulation
% when control is passed to an untrusted program.

% Many high-level languages guarantee that function function calls are
% well-bracketed. This can easily be achieved by having a trusted call
% stack. On a capability machine, it is possible to enforce well
% barcketedness without a trusted stack. When ensuring well-bracketed
% the two main challenges calls are 1) preventing storage of return
% pointers and 2) encapsulation of the local stack frame. If we don't
% have the first property, then an adversary can store the return
% capability in a call and jump to it in a later call. If the local
% stack frame is not properly encapsulated, then an adversary can break
% the well-bracketedness by jumping to a different calls return
% capability.

% In order to deal with the two challenges, we make sure that all return
% capabilities are local and that there are no global capabilities with
% permit write local permission. This makes sure that there is no way to
% save a return capability in persistent local state.\lau{this does not really make
% sense as the local stack frame can be seen as local state} It is,
% however, to limiting to not provide any means to store the return
% capabilities as a piece of code may need to store a return capability
% before jumping to an adversary. It is therefore needed to provide a
% local capability with permit write local permission which can be used
% as a stack. The stack should now be the only place one can store the
% return capability, but it should also be the only place one can store
% the stack capability.

% Using local return capabilities and putting a stack abstraction on a
% local capability permit write local capability is a good start, but as
% it turns out, it is not wnough. An adversary could do the following 
% \begin{enumerate}
% \item In a first invocation, fill the stack frame with copies of the
%   current stack capability.
% \item In a later call, the adversary may be able to load the stack
%   capability from the previous call.
% \item If we have used a larger part of the stack than in the first
%   call, then the adversary has access to part of our local stack frame.
% \end{enumerate}

% The above attack shows that it is tricky to ensure
% well-bracketedness. It raises the question whether it is possible to
% do so it is completely watertight. The attack utilized the stack, so
% it is necessary to have a trusted stack or can we do without? In
% general it raises the question, how do we reason about capability
% machines and local capabilities, so we are sure that what ever calling
% scheme we come up with actually provably ensures well-bracketed calls.

% In the following paper, we present the following contributions
% \begin{enumerate}
% \item 
% \end{enumerate}
